Optimal Parameter Selection in Markov State Models for Biomolecular Conformational Dynamics

### 2013-11-28 paper/paper2.tex
Final updates to the paper including a new abstract, fixed citations, and reformatting a couple sentences

\documentclass[twocolumn,floatfix,nofootinbib,aps]{revtex4-1}
\usepackage[utf8]{inputenc}

\usepackage{amsmath}    % need for subequations
\usepackage{amssymb}    % for symbols
\usepackage{graphicx}   % need for figures
\usepackage{verbatim}   % useful for program listings
\usepackage{color}      % use if color is used in text
\usepackage{subfigure}  % use for side-by-side figures
\usepackage{hyperref}   % use for hypertext links, including those to external
                        % documents and URLs
\usepackage[capitalise]{cleveref}   % use for referencing figures/equations
\usepackage{soul}
\usepackage{bm}

\begin{document}

\title{Optimal Parameter Selection in Markov State Models for Biomolecular Conformational Dynamics}
\author{Robert T. McGibbon}
\author{Christian R. Schwantes}
\author{Vijay S. Pande}

\begin{abstract}
Markov state models (MSMs) are a powerful tool for the analysis of molecular dynamics simulations, but have been hampered by the need for manual selection of the number of states. We report a new method for the optimal selection of the number of states in an MSM based on the Bayesian information criterion. We demonstrate the approach on three systems of increasing complexity...
\end{abstract}

\maketitle

\section{Introduction}
Protein are highly complex molecular machines, and their dynamics are an essential aspect of biomolecular function. These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, conformational change between functional native substates, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}. Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex multi-state kinetics \cite{Cosa2006Evidence, Zhang2011Direct}. But the atomic-resolution characterization of these dynamics is often an enormous challenge -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Mertens2010Structural, Tzeng2011Protein}.

Computer simulations can complement experiments by providing atomic insight into conformational dynamics. With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation data sets\cite{Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS, Buch2010High}. In fact, the minimally-biased kinetic analysis of such simulations is often a central bottleneck and presents a major challenge to the field. The analysis paradigms often entail the construction of lower resolution models parametrized from the high resolution simulation data set which capture the essential features in an interpretable framework\cite{Freddolino2010Challenges, Lane2013Milliseconds}. For example, by projecting the data down onto one or two degrees of freedom we create a simpler model for the system, such as one characterized by diffusion along a single reaction coordinate\cite{Best2010Coordinate}.

Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high and low-resolution models\cite{Pande2010Everything, Beauchamp2012Simple, Prinz2011Markov, Bowman2013Quantitative}. Such detailed models maintain quantitative agreement with the underlying simulation data, while low-resolution models capture the salient features of the potential energy landscape, sacrificing some degree of model complexity. In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states. The two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which parameterize the kinetics. The matrix of transition probabilities can be used to locate the systems transition paths\cite{Weinan2006Towards}, and its dominant eigenvectors to identify the metastable states\cite{Deuflhard2000Identification} and long-timescale dynamical modes\cite{}.

@@

\section{Prior Work}
In the context of MSMs, model selection has traditionally been performed by analyzing the self-consistency of the models with respect to changes in the Markov lag time.

\begin{itemize}
\item Swope-Pitera http://pubs.acs.org/doi/abs/10.1021/jp037421y
\item Noe http://www.pnas.org/content/106/45/19011
\item Park and Pande
\item Sergio
\item Kellogg
\end{itemize}

After diagonalization of the transition matrix, the relaxation of any initial probability distribution can be described by a superposition of exponential relaxation processes, with a characteristic time given by $-\frac{1}{\ln \lambda_i}$.

\section{Likelihood of a Markov State Model}
With the kinetic model expressed as a set of pairwise state to state transition probabilities at a given lag time, the likelihood of an ensemble of trajectories after projection into the indicator function basis is given simply by the product of the transition matrix elements along the observed trajectories. However, as we vary the number of states, it is not permissible to simply compare these likelihoods as part of an optimization of the state definitions. In doing so, the optimal model would always be the trivial one state model, whose computed likelihood is unity regardless of the data.

The appropriate likelihood is instead a path action in phase space, on which the discrete states are merely an indicator function basis. With $s(X)$ as the function mapping conformations into the indicator function basis set, and $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory, the likelihood can be written as:

\begin{equation}
\begin{split}
P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
\cdot \prod_{t=1}^{N} p_{s(\mathbf{x}_t)}(\mathbf{x}_{t}) \cdot d\mathbf{x}^N
\label{eq:like}
\end{split}
\end{equation}

With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state transition probabilities, $T(s_i \rightarrow s_j)$, and so-called emission distributions of each state, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $\mathbf{x}_t$ is within a certain state, $s(\mathbf{x}_t)$.

For example, consider two Markov state models sharing the same transition matrix, $T$. In one model, the state emission distributions are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the states. If an observed trajectory does go through the first models' regions of high likelihood, it is appropriately termed a more likely model given the data.

This emission distribution can be potentially modeled in multiple ways, each of which, in combination with a transition matrix, parameterizes a different statistical model. We propose two possible distributions: first, one may model the conditional distribution $P(\mathbf{x}_t | s(\mathbf{x}_t))$ as a uniform distribution over the (hyper)volume of $s(\mathbf{x}_t)$, as shown in eq. \ref{eq:like_vol}. This model has the advantage of appearing agnostic with respect to intra-state dynamics but is intractable in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation. A more tractable alternative is a Gaussian emission model where $P(\mathbf{x}_t | s(\mathbf{x}_t))$ is modeled as multivariate normal, where the pairwise overlap between the emission distributions is treated as a truncation error, as shown in eq.  \ref{eq:like_vol} and used in the popular X-means algorithm.\cite{Pelleg2000Xmeans} The truncation error steps from the fact that the emission probability distributions from every state assign nonzero probability to each conformation in a dataset, while the parameterization procedure for the Markov transition probabilities assumes each conformation to be uniquely assigned to a single state. Relaxing this approximation is possible with the forward-backward algorithm,\cite{Rabiner1989Tutorial} but is beyond the scope of this work.

%The simplest definition would be a uniform distribution since we are constructing an MSM that is agnostic to intra-state dynamics. When properly normalized, the uniform distribution takes a value of the reciprocal of the state's volume. This makes for an easy expression, however, volumes in high-dimensional states can be difficult to compute \ref{eq:like_vol}. For this reason, we can also choose other emission distributions that may be more tractable. In fact, \citet{Pelleg2000Xmeans} have used a gaussian emission distribution to define a likelihood to compare two k-means clusterings. The result of this emission distribution, is that the likelihood is proportional to the sum of square distances, which is precisely what is optimized in k-means clustering. If we think of k-centers as a poor man's version of k-means, then we can apply this same likelihood.

\begin{equation}
\label{eq:like_vol}
\begin{split}
P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
	\cdot \prod_{t=1}^N \frac{1}{V_{s(\mathbf{x}_{t})}} \cdot d\mathbf{x}^N
\end{split}
\end{equation}

\begin{equation}
\label{eq:like_mvn}
\begin{split}
&P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
	&\;\;\cdot \prod_{t=1}^N\Bigg[ \frac{1}{(2 \pi)^\frac{d}{2} |\Sigma_{s(\mathbf{x}_t)}|^\frac{1}{2}} \\
    &\;\;\cdot \exp\left(-\frac{1}{2} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})^T \Sigma_{s(\mathbf{x}_t)}^{-1} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})\right)\Bigg] \cdot d\mathbf{x}^N
\end{split}
\end{equation}

where $\bm{\mu}_{s(\mathbf{x}_t)}$ is the cluster center assigned to $\mathbf{x}_t$ and $\Sigma_{s(\mathbf{x}_t)}$ is the corresponding covariance matrix. For simplicity, we follow \citet{Pelleg2000Xmeans} and employ a spherical Gaussian model with a single shared variance parameter across the states estimated by maximum likelihood, under which conditions eq. \ref{eq:like_mvn} reduces to
\begin{equation}
\label{eq:like_mvn2}
\begin{split}
P\left[\{\mathbf{x}_t\}_{t=1}^{N}\right] & d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
	&\cdot \prod_{t=1}^N \left(2 \pi \hat{\sigma}^2\right)^{-\frac{d}{2}} \exp\left(-\frac{||\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)}||^2}{2 \hat{\sigma}^2}\right)
\end{split}
\end{equation} where
\begin{equation}
\label{eq:mle_sigma}
\hat{\sigma}^2 = \frac{1}{T - k} \sum_{i=0}^T || x_i - \mu_{s(x_i)} ||^2
\end{equation} where $k$ is the number of states.

\section{Cross Validation and the Bayes Information Criterion}

Likelihood maximization is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to overfitting\cite{Liddle2007Information, Hastie01Elements}. Statistical learning theory provides a number of alternative approaches for this problem. Conceptually, the most straightforward is a full Bayesian treatment in which all unknown model parameters are represented by probability distributions. The evidence for a model is computed by formally integrating over the model parameters and the evidence ratio, or Bayes factor\cite{Gelfand94Bayesian}, then provides a rigorous basis of model selection that appropriately punishes overly complex models as they become poorly constrained in parameter space. Unfortunately such approaches are intractable for problems of this size because of the need to integrate over all possible Markov models of a given size.

Instead, we explore three alternative procedures: cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC} for choosing the number of states in a Markov state model. In cross validation, we attempt to directly measure the Markov model's generalization error. We parameterize the model, building both the state space and transition matrix on a subset of the available molecular dynamics trajectories, but evaluate the likelihood on the left-out portion. The AIC and BIC are augmented likelihood function approaches which do not require leaving out portions of the available data during fitting, and instead employ asymptotic approximations to the Bayesian evidence to directly penalize models with additional free parameters.

\begin{equation}
\label{eq:bic}
\mathrm{BIC} \equiv -2\cdot \ln L + k \cdot \ln N
\end{equation}
\begin{equation}
\label{eq:eic}
\mathrm{AIC} \equiv -2\cdot \ln L + 2 \cdot k
\end{equation}

where $L$ is the likelihood, $k$ is the number of free parameters, and $N$ is the number of data points, assumed to be independent and identically distributed.

@@

\begin{figure}
\centering
\includegraphics[width=3in]{figs_final/mull_ww.png}
\caption{Systems studied in this work. (A) Langevin dynamics on the two dimensional M\"{u}ller potential. (b) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E. Shaw research \cite{Shaw2010Atomic}.}
\end{figure}

The uniform distribution emission model presents a computational challenge: its use requires the calculation of the (hyper)volume of the MSM's states, which, when defined by clustering are high-dimensional Voronoi cells. While trivial in two or three dimensions, this computational geometry task becomes challenging in high-dimensional settings. The computation of such volumes has occupied significant attention in recent years in the computational geometry literature, especially via randomized algorithms\cite{Kannan97, Simonovits03, Lovasz03}. We opt to approximate the volumes using naive Monte Carlo rejection sampling, which we find tractable for large systems only when the molecular dynamics dataset is first projected into a suitable small vector space of up to perhaps ten dimensions.

A further challenge is the procedure by which to model the volume of states which are at the edge of the MSM -- whose Voronoi cells extend to infinity in some direction. In these cases the Voronoi cells are of unbounded volume. Because the convex hull of our simulation datasets are computationally inaccessible, we define the outer extent of our data sets to be the set of all trial points such that the nearest sampled configuration to the trial point is closer than a certain cutoff, $R$.

For cross validation, we find it essential to use a dense prior during transition matrix estimation. Because the maximum likelihood transition matrix fit on a subset of the data assigns probability zero to unobserved transitions, we observe cross validation log likelihoods of negative infinity in all but the most data-rich regimes with maximum likelihood methods. We used a dirichlet prior on each entry of the transition matrix, which

For simplicity, we use the Jeffreys prior on each row of the transition matrix -- a Dirichlet prior with $\alpha=\frac{1}{2}$ -- which can be interpreted as a psuedocount of one half of an observed transition between each pair of states.

\section{Results and Discussion}
\subsection{M\"{u}ller Potential}
We simulated two trajectories with $10^6$ steps of Langevin dynamics on the M\"{u}ller potential\cite{}. Simulations were performed with a timestep $t=0.1$, friction coefficient of $\gamma=10^3$, and $kT=15$. The first trajectory was clustered using the $k$-centers clustering algorithm, and state volumes were computed for the uniform emission model using $10^5$ rounds of Monte Carlo rejection sampling with a padding distance of $R=0.2$ around the state centers from the 500 state model to define the dataset's envelope. The second trajectory was used as a test set.

@@

\begin{figure}[h]
\centering
\includegraphics[width=3in]{figs_final/mull_likes.png}
\caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states. There are many methods that can be used to test the transferability of an MSM. We've plotted the AIC, BIC, and cross-validation scores for the M\"uller potential simulations. These results indicate that, at least for this dataset, the BIC and AIC may penalize the number of states too heavily.}
\label{fig:mullerlike}
\end{figure}

As shown in \cref{fig:mullerlike}, models built with too few states achieve a drastically reduced likelihood, but above a threshold region the likelihood increases relatively slowly. Here, the AIC and BIC's penalty on the number of parameters, which scales with the square of the number of states, begins to dominate. The optimal models according to the three methods are between 200 and 1000 states for this system. This observation is consistent with the convergence of the relaxation timescales of the models.

Interestingly, the AIC and BIC penalize the larger state models much more than the cross-validation approach. This could be due to the simplicity of this system. Since the M\"uller potential is only two-dimensional, it is difficult to actually produce a drastically over-fit model. We suspect that the cross-validation score on a real protein system would be closer to the BIC and AIC. However, we note, that cross-validation requires estimating a model without some subset of the data, which is not feasible for large systems in the data-sparse regime.

>>>>

%\documentclass[twocolumn,floatfix,nofootinbib,acs]{revtex4-1}
\documentclass[journal=jpcbfk, layout=twocolumn, manuscript=article]{achemso}
\setkeys{acs}{articletitle = true}
\usepackage[utf8]{inputenc}

\usepackage{amsmath}    % need for subequations
\usepackage{amssymb}    % for symbols
\usepackage{graphicx}   % need for figures
\usepackage{verbatim}   % useful for program listings
\usepackage{color}      % use if color is used in text
\usepackage{subfigure}  % use for side-by-side figures
\usepackage{hyperref}   % use for hypertext links, including those to external
                        % documents and URLs
\usepackage[capitalise]{cleveref}   % use for referencing figures/equations
\usepackage{soul}
\usepackage{bm}
\usepackage{achemso}

\title{Optimal Parameter Selection in Markov State Models for Biomolecular Conformational Dynamics}
\author{Christian R. Schwantes}
\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
\altaffiliation{Contributed equally to this work}
\author{Robert T. McGibbon}
\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
\altaffiliation{Contributed equally to this work}
\author{Vijay S. Pande}
\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
\alsoaffiliation{Biophysics Program, Stanford University, Stanford CA 94305, USA}
\alsoaffiliation{Department of Computer Science, Stanford University, Stanford CA 94305, USA}
\alsoaffiliation{Department of Structural Biology, Stanford University, Stanford CA 94305, USA}
\email{pande@stanford.edu}

\mciteErrorOnUnknownfalse

\begin{document}
\maketitle
\begin{abstract}
Markov state models provide a powerful framework for the analysis of biomolecular conformation dynamics in terms of their metastable states and transition rates. These models provide both a quantitative and comprehensible description of the long-timescale dynamics of large molecular dynamics with a Master equation, and have been successfully used to study protein folding, protein conformational change, and protein-ligand binding. However, to achieve satisfactory performance, existing methodologies often require expert intervention when defining the model's discrete state space. Here, we show that by consideration of the states' conditional distribution over conformations, the state space can be evaluated and optimized using the Markov state model's likelihood function and standard statistical model selection tools. Application of these techniques on two 100$\mu s$ molecular dynamics trajectories of the Fip35 WW domain shows agreement with existing techniques based on self-consistency of the model's relaxation timescales, with more suitable results in regimes in which those techniques encourage over-fitting. By removing the need for expert tuning, these methods should reduce modeling bias and lower the barriers to entry in Markvov state model construction.
\end{abstract}

\section{Introduction}
Protein are highly complex molecular machines, and their dynamics are an essential aspect of biomolecular function. These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, conformational change between functional native sub-states, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}. Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex multi-state kinetics \cite{Cosa2006Evidence, Zhang2011Direct}. But the atomic-resolution characterization of these dynamics is often an enormous challenge -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Mertens2010Structural, Tzeng2011Protein}.

Computer simulations can complement experiments by providing atomic insight into conformational dynamics. With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation ensembles\cite{Gotz2012Routine, Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS, Buch2010High}. In fact, the minimally-biased kinetic analysis of such simulations is often a central bottleneck and presents a major challenge to the field. The analysis paradigms often entail the construction of lower resolution models parametrized from the high resolution simulation data set which capture the essential features in an interpretable framework\cite{Freddolino2010Challenges, Lane2013Milliseconds}. For example, by projecting the data down onto one or two degrees of freedom we create a simpler model for the system, such as one characterized by diffusion along a single reaction coordinate\cite{Best2010Coordinate}.

Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high and low-resolution models\cite{Chodera2007Automatic, Prinz2011Markov, Beauchamp2012Simple, Bowman2013Quantitative}. Such detailed models maintain quantitative agreement with the underlying simulation data, while low-resolution models capture the salient features of the potential energy landscape, sacrificing some degree of model complexity. In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states. The two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which parameterize the kinetics. The matrix of transition probabilities can be used to locate the systems transition paths\cite{Weinan2006Towards}, and its dominant eigenvectors to identify the metastable states\cite{Deuflhard2000Identification} and long-timescale dynamical modes\cite{}.

@@

\section{Prior Work}
%After diagonalization of the transition matrix, the relaxation of any initial probability distribution can be described by a superposition of exponential relaxation processes, with a characteristic time given by $-\frac{1}{\ln \lambda_i}$.
In the context of MSMs, model selection has traditionally been performed by analyzing the self-consistency of the models with respect to changes in the Markov lag time. As suggested by \citet{Swope2004Describing}, given a Markov model at a lag time of $\tau$, $T(\tau)$, we expect that a model with a lag time of $n\tau$ should merely be the $n^\textnormal{th}$ power of $T(\tau)$:
$$ T(n\tau) = T(\tau)^n $$ As such, the eigenvalues, and correspondingly the relaxation timescales, should remain constant as the lag time increases. The resulting test of a model's quality consists of calculating the eigenspectra of models built at many lag times and choosing models whose eigenvalues have converged.

In practice, while this test has been useful for many \cite{Noe2009Constructing}, it has some flaws. In particular, it is generally observed that most models' timescales do not converge, but steadily increase, making it is difficult to judge whether a particular model passes the test and encouraging the use of ever-more states. Additionally, it is difficult to compare models built on two state decompositions, as neither may have convergent timescales. Often, the choice is made to use the model whose timescales better match an experimental observable.

Other evaluation strategies include a test based on comparison to a second-order Markov model \cite{Park2006Validation}. Additionally, Bayesian methods that involve marginalizing over all possible MSMs have been developed \cite{Bacallado2009Bayesian, *Bacallado2011Bayesian}. However, these methods are often prohibitively expensive on larger data sets. In contrast, \citet{Kellogg2012Evaluation} proposed a tractable method for comparing MSMs built with different state decompositions, based on a likelihood function for an MD trajectory given an MSM, discussed below.

\section{Likelihood of a Markov State Model}
Bayes' rule provides a foundation for model selection, by establishing the proportionality of the probability of a model given the data to the probability of the data given the model. Without a strong prior, model selection is reduced to the search for the model that maximizes the likelihood of the data.

Thus we seek to formulate this likelihood function for Markov state models. With the kinetic model expressed as a set of pairwise state to state transition probabilities at a given lag time, the likelihood of an ensemble of trajectories after projection into the MSM state space is given simply by the product of the transition matrix elements along the observed trajectories. However, as we vary the number of states, it is not permissible to simply compare these likelihoods as part of an optimization of the state definitions. In doing so, the optimal model would always be the trivial one state model, whose computed likelihood is unity regardless of the data.

The appropriate likelihood is instead a path action in phase space, on which the discrete states are merely an indicator function basis. With $s(\mathbf{x})$ as the function mapping conformations into the indicator function basis set, and $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory, the likelihood can be written as:

\begin{equation}
\begin{split}
P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
\cdot \prod_{t=1}^{N} P(\mathbf{x}_t | s(\mathbf{x}_t)) \cdot d\mathbf{x}^N
\label{eq:like}
\end{split}
\end{equation}

With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state Markov transition probabilities, $T(s_i \rightarrow s_j) \equiv P(S_{t+1}=s_j | S_{t}=s_i)$, and so-called emission distributions of each state, $P(\mathbf{x} | s(\mathbf{x}))$, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $\mathbf{x}$, is within a certain state, $s_i$. By convention, we define $P(\mathbf{x}| s(\mathbf{x}))$ to be zero for all $\mathbf{x}$ such that $s(\mathbf{x}) \neq s_i$.

This emission distribution can be potentially modeled in multiple ways, each of which, in combination with a transition matrix, parametrizes a different statistical model. \citet{Kellogg2012Evaluation} used a discriminative model with:
\begin{equation}
P(\mathbf{x}_t | s(\mathbf{x}_t)) = \big| \{ \mathbf{x}_k : s(\mathbf{x}_k) = s(\mathbf{x}_t) \} \big|^{-1}
\end{equation} This model has the training data set as its support rather than phase space, which makes it unable to generalize and assign probability to new data. In certain circumstances, this is an undesirable property. For example, protocols that involve fitting and validating models on separate data sets (e.g. cross-validation) are impossible when the statistical model lacks the capacity to describe new data.

Instead, we seek a generative model whose emission distributions are supported on phase space (i.e. standard distributions over $\mathbb{R}^{3N_{atoms}}$). We propose two possible emission distributions, $P(\mathbf{x} | s(\mathbf{x}))$: first, a uniform distribution over the (hyper)volume of each state, as shown in \cref{eq:like_vol}. This model has the advantage of appearing agnostic with respect to intra-state dynamics but is intractable in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation.

%For example, consider two Markov state models sharing the same transition matrix, $T$. In one model, the state emission distributions are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the states. If an observed trajectory does go through the first models' regions of high likelihood, it is appropriately termed a more likely model given the data.

%The simplest definition would be a uniform distribution since we are constructing an MSM that is agnostic to intra-state dynamics. When properly normalized, the uniform distribution takes a value of the reciprocal of the state's volume. This makes for an easy expression, however, volumes in high-dimensional states can be difficult to compute \ref{eq:like_vol}. For this reason, we can also choose other emission distributions that may be more tractable. In fact, \citet{Pelleg2000Xmeans} have used a gaussian emission distribution to define a likelihood to compare two k-means clusterings. The result of this emission distribution, is that the likelihood is proportional to the sum of square distances, which is precisely what is optimized in k-means clustering. If we think of k-centers as a poor man's version of k-means, then we can apply this same likelihood.
\begin{equation}
\label{eq:like_vol}
\begin{split}
%P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
 P(\mathbf{x}_t | s(\mathbf{x}_t)) = \frac{1}{V_{s(\mathbf{x}_{t})}}
\end{split}
\end{equation}

A more tractable alternative is a Gaussian emission model where $P(\mathbf{x} | s(\mathbf{x}))$ is modeled as multivariate normal \cref{eq:like_mvn}.
\begin{equation}
\label{eq:like_mvn}
\begin{split}
P(\mathbf{x}_t | s(\mathbf{x}_t)) = \Bigg[ \frac{1}{(2 \pi)^\frac{d}{2} |\Sigma_{s(\mathbf{x}_t)}|^\frac{1}{2}} \\
   \exp\left(-\frac{1}{2} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})^T \Sigma_{s(\mathbf{x}_t)}^{-1} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})\right)\Bigg] 
\end{split}
\end{equation} where $\bm{\mu}_{s(\mathbf{x}_t)}$ is the cluster center assigned to $\mathbf{x}_t$ and $\Sigma_{s(\mathbf{x}_t)}$ is the corresponding covariance matrix. For simplicity, we follow \citet{Pelleg2000Xmeans} and employ a spherical Gaussian model with a single shared variance parameter across the states estimated by maximum likelihood, under which it reduces to:
\begin{equation}
\label{eq:like_mvn2}
 P(\mathbf{x}_t | s(\mathbf{x}_t)) = \frac{1}{\left(2 \pi \hat{\sigma}^2\right)^{\frac{d}{2}}} \exp\left(-\frac{||\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)}||^2}{2 \hat{\sigma}^2} \right)
\end{equation} where
\begin{equation}
\label{eq:mle_sigma}
\hat{\sigma}^2 = \frac{1}{N - k} \sum_{t=1}^N || \mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)} ||^2
\end{equation} and $k$ is the number of states.

Since the likelihood requires the emission distributions to have zero overlap -- such that each conformation only has nonzero emission probability from a single state, the nonzero overlap between the Gaussian emission distributions introduces a truncation error. One approach to ameliorate this issue is to truncate the density functions at the state boundaries, and renormalize. We assume that the overlaps are small, and use the density defined in \cref{eq:like_mvn2}. An alternative approach for relaxing this approximation is possible with the forward-backward algorithm,\cite{Rabiner1989Tutorial} but is beyond the scope of this work.

\section{Cross Validation and the Bayes Information Criterion}
Likelihood maximization is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to over-fitting\cite{Liddle2007Information, Hastie01Elements}. Statistical learning theory provides a number of alternative approaches for this problem. Conceptually, the most straightforward is a full Bayesian treatment in which all unknown model parameters are represented by probability distributions. The evidence for a model is computed by formally integrating over the model parameters and the evidence ratio, or Bayes factor\cite{Gelfand94Bayesian}, then provides a rigorous basis of model selection that appropriately punishes overly complex models as they become poorly constrained in parameter space. Unfortunately such approaches are intractable for problems of this size because of the need to integrate over all possible Markov models of a given size.

Instead, we explore three alternative procedures for choosing the number of states in an MSM: cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC}. Cross-validation attempts to directly measure the Markov model's generalization error. First, the model is parameterized by building both the state space and transition matrix on a subset of the available molecular dynamics trajectories, then the likelihood is evaluated on the left-out portion. This scheme can be repeated many times with different partitions of the data set. The AIC and BIC are augmented likelihood function approaches which do not require leaving out portions of the available data during fitting, and instead employ asymptotic approximations to the Bayesian evidence to directly penalize models with additional free parameters.
\begin{equation}
\label{eq:bic}
\mathrm{BIC} \equiv -2\cdot \ln L + \kappa \cdot \ln N
\end{equation}
\begin{equation}
\label{eq:aic}
\mathrm{AIC} \equiv -2\cdot \ln L + 2 \cdot \kappa
\end{equation} where $L$ is the likelihood, $\kappa$ is the number of free parameters, and $N$ is the number of data points, assumed to be independent and identically distributed. Model selection is performed by a minimization of the criterion.

@@

\begin{figure}
\centering
\includegraphics[width=3in]{figs_final/mull_ww.png}
\caption{Systems studied in this work. (A) Langevin dynamics on the two dimensional M\"{u}ller potential \cite{Muller1980Reaction}. (b) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E. Shaw research \cite{Shaw2010Atomic}.}
\end{figure}

The uniform distribution emission model presents a computational challenge: its use requires the calculation of the (hyper)volume of the MSM's states, which, when defined by clustering, are high-dimensional Voronoi cells. While trivial in two or three dimensions, this computational geometry task becomes challenging in high-dimensional settings. The computation of such volumes has occupied significant attention in recent years in the computational geometry literature, especially via randomized algorithms\cite{Kannan97, Simonovits03, Lovasz03}. We opt to approximate the volumes using naive Monte Carlo rejection sampling, which we find tractable for large systems only when the molecular dynamics data set is first projected into a suitable small vector space of up to perhaps ten dimensions.

A further challenge is the description of the states that are at the edge of the MSM -- whose Voronoi cells extend to infinity in at least one direction. In these cases, the Voronoi cells are of unbounded volume. Instead we wish to truncate these states by bounding them by the convex hull of the data set. Because the convex hull of our simulation data sets are computationally inaccessible, we define the outer extent of our data sets to be the set of all trial points such that the nearest sampled conformation to the trial point is closer than a certain cutoff, $R$.

For cross validation, we find it essential to use a dense prior during transition matrix estimation. Because the maximum likelihood transition matrix fit on a subset of the data assigns probability zero to unobserved transitions, we observe cross validation log likelihoods of negative infinity in all but the most data-rich regimes. We used a Dirichlet prior with a pseudo-count of $\frac{1}{k}$ for each possible transition, where $k$ is the number of states in the model.

\section{Results and Discussion}
\subsection{M\"{u}ller Potential}
We simulated two trajectories with $10^6$ steps of Langevin dynamics on the M\"{u}ller potential\cite{Muller1980Reaction}. Simulations were performed with a timestep $t=0.1$, friction coefficient of $\gamma=10^3$, and $kT=15$. The first trajectory was clustered using the $k$-centers clustering algorithm, and state volumes were computed for the uniform emission model using $10^5$ rounds of Monte Carlo rejection sampling with a cutoff of $R=0.28$ from the 50 state model's centroids. The second trajectory was used as a test set. All MSMs were built using a lag time of $30 steps$.

@@

\begin{figure}[h]
\centering
\includegraphics[width=3in]{figs_final/mull_likes.png}
\caption{We calculated the training and test set log-likelihoods and model selection criteria for models built with 50-1000 states using both the uniform ({\bf A}) and multivariate normal ({\bf B}) emission distributions. Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray). For both likelihoods, the BIC penalizes complexity more than the AIC and test set log-likelihood. The optimal window is consistent with the convergence of the relaxation timescales (right frames).}
\label{fig:mullerlike}
\end{figure}

As shown in \cref{fig:mullerlike}, models built with too few states achieve a drastically reduced likelihood, but above a threshold region the likelihood increases relatively slowly. The penalty on the number of parameters in \cref{eq:bic} and \cref{eq:aic} begins to dominate. The optimal models, according to the three methods are between 200 and 600 states for this system, which is consistent with the convergence of the relaxation timescales of the models.

The AIC and BIC penalize the larger state models much more than the test set log-likelihood. On the two-dimensional potential, in our data-rich regime it is difficult to produce an over-fit model. Cross-validation requires fitting multiple models on subsets of the data, and so is not feasible for larger systems in the data-poor regime.
